---
title: "CoGAPS Report | `r params$sample` | `r params$npatterns` patterns"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
    code-tools: true
    highlight-style: github
    embed-resources: true
    anchor-sections: true
    smooth-scroll: true
lightbox: true
execute:
  cache: false
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
params:
  sample: "SAMPLE"
  npatterns: 0
  outdir: ""
  contrib_cells: ""
  contrib_genes: ""
  markers_excel: ""
  enrich_csv: ""
  overr_csv: ""
  kw_csv: ""
  pairwise_csv: ""
  dotplot: ""
  png_specificity_patterns: ""
  png_specificity_metagroups: ""
  png_enrich_unique: ""
  png_overr_unique: ""
  png_enrich_heatmap: ""
  png_overr_heatmap: ""
  png_enrich_heatmap_manually: ""
  png_overr_heatmap_manually: ""
  umap_nmf: ""
  umap_nb_cells: ""
  weight_nmf: ""
  benchmark_tsv: ""
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
    library(readxl)
    library(readr)
    library(dplyr)
    library(knitr)
    library(kableExtra)
})
s <- params$sample
n <- params$npatterns
```

## Overview

```{r tbl-params}
n_learned <- NA
n_cells <- NA

if (file.exists(params$contrib_cells) && params$contrib_cells != "") {
  contrib <- read_tsv(params$contrib_cells, show_col_types = FALSE)
  n_learned <- sum(grepl("^Pattern", colnames(contrib)))
  n_cells <- nrow(contrib)
}

data.frame(
    Parameter = c(
      "Sample",
      "Number of targeted patterns",
      "Number of learned patterns",
      "Number of cells"
    ),
    Value = c(
      s,
      as.character(n),
      as.character(n_learned),
      as.character(n_cells)
    )
) |>
  kable(caption = "Sample information") |>
  kable_styling(full_width = FALSE)
```

```{r tbl-benchmark}
if (file.exists(params$benchmark_tsv) && params$benchmark_tsv != "") {

  read_tsv(params$benchmark_tsv, show_col_types = FALSE) |>
    kable(digits = 2, caption = "CoGAPS benchmark") |>
    kable_styling(bootstrap_options = c("striped", "hover"))

}
```

## Contributions

::: {.panel-tabset}

### Cells

```{r tbl-contrib-cells}
read_tsv(params$contrib_cells, show_col_types = FALSE) |>
    head(50) |>
    kable(caption = "P matrix (sampleFactors)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")

knitr::asis_output("<p style='margin-top: 8px;'><em>Note:</em> Only the first 50 rows are shown.</p>")
```

### Genes

```{r tbl-contrib-genes}
read_tsv(params$contrib_genes, show_col_types = FALSE) |>
    head(50) |>
    kable(caption = "A matrix (featureLoadings)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")

knitr::asis_output("<p style='margin-top: 8px;'><em>Note:</em> Only the first 50 rows are shown.</p>")
```

:::

## UMAP — Pattern weight

```{r fig-weight-nmf}
#| out-width: 100%
include_graphics(params$weight_nmf)
```

## UMAP — Dominant pattern

::: {.panel-tabset}

### NMF

```{r fig-umap-nmf}
#| out-width: 100%
include_graphics(params$umap_nmf)
```

### Cell counts

```{r fig-nb-cells}
#| out-width: 100%
include_graphics(params$umap_nb_cells)
```

:::

## Pattern specificity

::: {.panel-tabset}

### By pattern

```{r fig-specificity-patterns}
#| fig-cap: "Pattern specificity"
#| out-width: 100%
include_graphics(params$png_specificity_patterns)
```

### By metagroup

```{r fig-specificity-metagroups}
#| fig-cap: "Metagroup specificity"
#| out-width: 100%
include_graphics(params$png_specificity_metagroups)
```

:::

## Statistics

::: {.panel-tabset}

### Kruskal-Wallis

```{r tbl-kw}
read_csv(params$kw_csv, show_col_types = FALSE) |>
    kable() |> kable_styling(bootstrap_options = c("striped", "hover"))
```

### Pairwise comparisons

```{r tbl-pairwise}
read_csv(params$pairwise_csv, show_col_types = FALSE) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")
```

:::

## Markers

::: {.panel-tabset}

### Dotplot

```{r fig-dotplot}
#| fig-cap: "Marker genes dotplot by fixing dominant pattern"
#| out-width: 100%
include_graphics(params$dotplot)
```

:::

## Hallmarks

::: {.panel-tabset}

### Enrichment — heatmap (get_function)

```{r fig-enrich-heatmap}
#| fig-cap: "Hallmarks enrichment heatmap"
#| out-width: 100%
include_graphics(params$png_enrich_heatmap)
```

### Enrichment — heatmap (manually)

```{r fig-enrich-heatmap-manually}
#| fig-cap: "Hallmarks enrichment heatmap (manually)"
#| out-width: 100%
include_graphics(params$png_enrich_heatmap_manually)
```

### Enrichment — unique

```{r fig-enrich-unique}
#| fig-cap: "Hallmarks enrichment (unique)"
#| out-width: 100%
include_graphics(params$png_enrich_unique)
```

### Overrepresentation — heatmap (get_function)

```{r fig-overr-heatmap}
#| fig-cap: "Hallmarks overrepresentation heatmap"
#| out-width: 100%
include_graphics(params$png_overr_heatmap)
```

### Overrepresentation — heatmap (manually)

```{r fig-overr-heatmap-manually}
#| fig-cap: "Hallmarks overrepresentation heatmap (manually)"
#| out-width: 100%
include_graphics(params$png_overr_heatmap_manually)
```

### Overrepresentation — unique

```{r fig-overr-unique}
#| fig-cap: "Hallmarks overrepresentation (unique)"
#| out-width: 100%
include_graphics(params$png_overr_unique)
```

### Tables

```{r tbl-enrich}
read_csv(params$enrich_csv, show_col_types = FALSE) |>
    select(-leadingEdge) |>
    kable(caption = "Hallmarks enrichment") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "350px")
```

```{r tbl-overr}
read_csv(params$overr_csv, show_col_types = FALSE) |>
    kable(caption = "Hallmarks overrepresentation") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "350px")
```

:::
