---
title: "CoGAPS Report | `r params$sample` | `r params$npatterns` patterns"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    code-fold: true
    code-tools: true
    highlight-style: github
    embed-resources: true
    anchor-sections: true
    smooth-scroll: true
lightbox: true
execute:
  cache: false
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
params:
  sample: "SAMPLE"
  npatterns: 0
  outdir: ""
  contrib_cells: ""
  contrib_genes: ""
  markers_excel: ""
  enrich_csv: ""
  overr_csv: ""
  kw_csv: ""
  pairwise_csv: ""
  dotplot: ""
  png_specificity_patterns: ""
  png_specificity_metagroups: ""
  png_enrich_unique: ""
  png_overr_unique: ""
  png_enrich_heatmap: ""
  png_enrich_heatmap_manually: ""
  png_overr_heatmap_manually: ""
  png_overr_heatmap: ""
  umap_nmf: ""
  umap_nb_cells: ""
  weight_nmf: ""
  benchmark_tsv: ""
  heatmap_score_pg: ""
  heatmap_score_pc_nmf: ""
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
    library(readxl)
    library(readr)
    library(dplyr)
    library(knitr)
    library(kableExtra)
})
s <- params$sample
n <- params$npatterns
```

## Overview

```{r tbl-params}
n_learned <- NA
n_cells <- NA

if (file.exists(params$contrib_cells) && params$contrib_cells != "") {
  contrib <- read_tsv(params$contrib_cells, show_col_types = FALSE)
  n_learned <- sum(grepl("^Pattern", colnames(contrib)))
  n_cells <- nrow(contrib)
}

data.frame(
    Parameter = c(
      "Sample",
      "Number of targeted patterns",
      "Number of learned patterns",
      "Number of cells"
    ),
    Value = c(
      s,
      as.character(n),
      as.character(n_learned),
      as.character(n_cells)
    )
) |>
  kable(caption = "Sample information") |>
  kable_styling(full_width = FALSE)
```

```{r tbl-benchmark}
if (file.exists(params$benchmark_tsv) && params$benchmark_tsv != "") {

  read_tsv(params$benchmark_tsv, show_col_types = FALSE) |>
    kable(digits = 2, caption = "CoGAPS benchmark") |>
    kable_styling(bootstrap_options = c("striped", "hover"))

}
knitr::asis_output("Allocated resources on GLICID: threads: 65 and mem_mb: 100 GB")
```

## Contributions

::: {.panel-tabset}

### Cells
```{r tbl-contrib-cells}
read_tsv(params$contrib_cells, show_col_types = FALSE) |>
    head(50) |>
    kable(caption = "P matrix (sampleFactors)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")

knitr::asis_output("
<p style='margin-top: 8px;'><em>Note:</em> Only the first 50 rows are shown.</p>
<p><em>Note:</em> The <code>dominant_pattern</code> column was added by assigning each cell to the pattern with the highest score across all patterns in the P matrix.</p>
")
```

### Genes

```{r tbl-contrib-genes}
read_tsv(params$contrib_genes, show_col_types = FALSE) |>
    head(50) |>
    kable(caption = "A matrix (featureLoadings)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")

knitr::asis_output("<p style='margin-top: 8px;'><em>Note:</em> Only the first 50 rows are shown.</p>")
```

:::

## UMAP — Pattern amplitudes
UMAP embedding of cells computed from NMF pattern contribution scores (P matrix), using uwot (cosine distance metric)
```{r fig-weight-nmf}
#| fig-cap: "NMF weight UMAP (based on P matrix)"
#| out-width: 100%
include_graphics(params$weight_nmf)
```

## Score 

```{r fig-heatmap_score}
#| fig-cap: "Heatmap of mean pattern activity across conditions"
#| out-width: 100%
include_graphics(params$heatmap_score_pg)
```
```{r fig-heatmap_score2}
#| fig-cap: "Heatmap of mean pattern activity across NMF clusters"
#| out-width: 100%
include_graphics(params$heatmap_score_pc_nmf)
```

## UMAP — Dominant pattern

::: {.panel-tabset}

### NMF
UMAP embedding of cells computed from NMF pattern contribution scores (P matrix), using uwot (cosine distance metric)
```{r fig-umap-nmf}
#| fig-cap: "NMF dominant UMAP"
#| out-width: 100%
include_graphics(params$umap_nmf)
```

### Cell counts

```{r fig-nb-cells}
#| out-width: 100%
include_graphics(params$umap_nb_cells)
```

:::

## Pattern specificity

::: {.panel-tabset}

### By pattern

```{r fig-specificity-patterns}
#| fig-cap: "Pattern specificity"
#| out-width: 100%
include_graphics(params$png_specificity_patterns)
```

### By metagroup

```{r fig-specificity-metagroups}
#| fig-cap: "Metagroup specificity"
#| out-width: 100%
include_graphics(params$png_specificity_metagroups)
```

:::

## Statistics

::: {.panel-tabset}

### Kruskal-Wallis
Global non-parametric test assessing whether NMF pattern weights differ across subclones.
```{r tbl-kw}
read_csv(params$kw_csv, show_col_types = FALSE) |>
    kable() |> kable_styling(bootstrap_options = c("striped", "hover"))
```

### Pairwise comparisons
Pairwise comparisons between subclones in each patterns performed using Wilcoxon tests with BH-adjusted p-values.
```{r tbl-pairwise}
read_csv(params$pairwise_csv, show_col_types = FALSE) |>
    kable() |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "400px")
```

:::

## Markers

::: {.panel-tabset}

### Table
The marker genes were defined using `patternMarker(threshold = "cut")` from **CoGAPS**.
```{r tbl}
df <- readxl::read_excel(params$markers_excel)
df |>
  head(50) |>
  knitr::kable(caption = "Top 50 Marker genes per pattern") |>
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover")) |>
  kableExtra::scroll_box(height = "400px")
```

### Dotplot
```{r fig-dotplot}
#| fig-cap: "Marker genes (top 5) by fixing dominant pattern"
#| out-width: 100%
include_graphics(params$dotplot)
```

:::

## Functional Annotation (Hallmarks)

::: {.panel-tabset}

### Enrichment

::: {.panel-tabset}

#### Heatmap (get_function)

Gene set enrichment was assessed using a ranking-based enrichment approach implemented with `getPatternGeneSet(method = "enrichment")` from CoGAPS, which applies `fgsea::gsea` to test enrichment of genes ranked by pattern amplitude against MSigDB Hallmark gene sets (retrieved via msigdbr).

```{r fig-enrich-heatmap}
#| fig-cap: "Hallmarks enrichment heatmap"
#| out-width: 100%
include_graphics(params$png_enrich_heatmap)
```

#### Heatmap (manually)

The enrichment analysis was performed with the `GSEA()` function from **ClusterProfiler**, with msigdbr hallmarks.

```{r fig-enrich-heatmap-manually}
#| fig-cap: "Hallmarks enrichment heatmap (manually)"
#| out-width: 100%
include_graphics(params$png_enrich_heatmap_manually)
```

#### Unique

```{r fig-enrich-unique}
#| fig-cap: "Hallmarks enrichment (unique)"
#| out-width: 100%
include_graphics(params$png_enrich_unique)
```

#### Table

```{r tbl-enrich}
read_csv(params$enrich_csv, show_col_types = FALSE) |>
    select(-leadingEdge) |>
    kable(caption = "Hallmarks enrichment (get_function)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "350px")
```

:::

### Over-representation

::: {.panel-tabset}

#### Heatmap (get_function)

Gene set enrichment was assessed using an over-representation analysis (ORA) implemented with `getPatternGeneSet(method = "overrepresentation")` from CoGAPS, which applies `fgsea::fora` to test enrichment of pattern marker genes against MSigDB Hallmark gene sets (retrieved via msigdbr).

```{r fig-overr-heatmap}
#| fig-cap: "Hallmarks overrepresentation heatmap"
#| out-width: 100%
include_graphics(params$png_overr_heatmap)
```

#### Heatmap (manually)

The analysis was performed with the `enricher` function from **ClusterProfiler**, with msigdbr hallmarks.

```{r fig-overr-heatmap-manually}
#| fig-cap: "Hallmarks overrepresentation heatmap (manually)"
#| out-width: 100%
include_graphics(params$png_overr_heatmap_manually)
```

#### Unique

```{r fig-overr-unique}
#| fig-cap: "Hallmarks overrepresentation (unique)"
#| out-width: 100%
include_graphics(params$png_overr_unique)
```

#### Table

```{r tbl-overr}
read_csv(params$overr_csv, show_col_types = FALSE) |>
    kable(caption = "Hallmarks overrepresentation (get_function)") |>
    kable_styling(bootstrap_options = c("striped", "hover")) |>
    scroll_box(height = "350px")
```

:::

:::
